MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-27 Manuscript/Manuscript-3.tex
vsn 3

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. We conducted a literature
survey in Google.scholar and found that of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\), 40\% interpreted
the strength of phylogenetic signal for at least one phenotypic trait.
Additionally, nearly 30\% of the 421 \(\lambda\) estimates between 0.25
and 0.75 were assigned a strength of signal, and because nearly half of
the 1,572 values reported were near the limits of the parameter (Figure
1), this percentage is even higher, as biological interpretation of
phylogenetic signal at the limits of \(\lambda\) are known. \hfill\break

@@

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that while PGLS parameters (e.g., \(\beta\)) are accurately estimated
with the inclusion of phylogenetic signal, estimates of \(\lambda\) are
not. We also find that estimates of \(\lambda\) vary widely for a given
input value of phylogenetic signal, and that the precision in estimating
\(\lambda\) is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. Additionally, the same \(\lambda_{est}\) may be obtained from
datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable estimate of the strength of
phylogenetic signal in phenotypic data. We subsequently derive a
standardized effect size for measuring the strength of phylogenetic
signal in phenotypic datasets, and apply the concept to two common
measures of phylogenetic signal: \(\lambda\) and \(Kappa\). Through
simulations across a wide range of conditions, we find that the
precision of effect sizes based on \(\lambda\) (\(Z_{\lambda}\)) are
less reliable than that those based on \(Kappa\) (\(Z_K\)), implying
that \(Z_K\) is a more robust effect size measure. Additionally, we
propose a two-sample test statistic that may be used to compare the
strength of phylogenetic signal among datasets, and provide an empirical
example to demonstrate its use. We conclude that estimates of
phylogenetic signal using Pagel's \(\lambda\) are often inaccurate, and
thus interpreting strength of phylogenetic signal in phenotypic datasets
based on this measure is compromised. By contrast, effect sizes obtained
from \(Kappa\) hold promise for characterizing phylogenetic signal, and
for comparing the strength of phylogenetic signal across datasets.

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal, with the worst
precision at intermediate levels of signal (\(\lambda_{in}\approx0.5\)),
and improved precision as input levels approached the extremes of its
range (i.e., \(\lambda_{in}\rightarrow0\) \&
\(\lambda_{in}\rightarrow1\)). Thus, estimates of \(\lambda\) were least
reflective of the true input signal at intermediate values.
Additionally, even at large levels of species richness, we found that
the range of \(\lambda_{est}\) still encompassed a substantial portion
of possible values (e.g., \(n=512\); \(\lambda_{in}=0.5\):
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\);
\(\lambda_{in} = 0.25 \to 0.65\)). Results were similar when \(\lambda\)
was co-estimated with regression parameters in PGLS regression (Figure
3). Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell 2010 (2010) (see Supporting
Information). However, estimates of phylogenetic signal were not, and
the spread of \(\lambda_{est}\) was even broader than that observed when
\(\lambda\) was estimated for only the dependent variable. Taken
together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

Adams and Collyer (2019b) recently suggested that the strength of
phylogenetic signal could be represented as an effect size, based on the
\(Kappa\) statistic and its empirical sampling distribution from
permutation. Here we formalize that suggestion, and find an effect size
as:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood. Note also that under the
null hypothesis, \(E(\lambda)=0\), a no phylogenetic signal is expected
under this condition (Freckleton et al. 2002). \hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4. Here two
things are evident. First, estimates of \(Z_K\) track the input
phylogenetic signal in a more linear fashion than do estimates of
\(Z_\lambda\). Second, the precision of \(Z_K\) is considerably more
stable as compared with \(Z_\lambda\), as coefficients of variation for
the set of \(\sigma^2_{Z_K}\) across input levels of phylogenetic signal
were an order of magnitude smaller for than was observed for
\(\sigma^2_{Z_\lambda}\) (Figure 4). This implied that estimates of the
strength of phylogenetic signal were more reliable and robust when using
\(Z_K\) as compared with \(Z_\lambda\). \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), it may be of interest to compare such measures across datasets.
This is useful, for instance, to determine whether the strength of
phylogenetic signal is greater in one phenotypic trait as compared with
another. As with other effect sizes derived from permutation
distributions (e.g., Adams and Collyer 2016, 2019a), a two-sample test
statistic may be found as:

@@

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above for
equation 2. Estimates of significance of \(\hat{Z}_{12}\) may be
obtained from a standard normal distribution. As with other two-sample
tests, \(\hat{Z}_{12}\) is typically considered a two-tailed test,
however directional (one-tailed) tests may be specified should the
empirical situation require it (see Adams and Collyer 2016, 2019a).

>>>>

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. We conducted a literature
survey in Google.scholar and found that of the 204 papers published in
2019 that estimated and reported Pagel's \(\lambda\), 40\% interpreted
the strength of phylogenetic signal for at least one phenotypic trait.
Additionally, because nearly half of the 1,572 \(\lambda\) values
reported were near the limits of the parameter (Figure 1), this
percentage is even higher, as biological interpretation of phylogenetic
signal at the limits of \(\lambda\) are known. \hfill\break

@@

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that while PGLS parameters (e.g., \(\beta\)) are accurately estimated
with the inclusion of phylogenetic signal, estimates of \(\lambda\) are
not. We also find that estimates of \(\lambda\) vary widely for a given
input value of phylogenetic signal, and that the precision in estimating
\(\lambda\) is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. Additionally, the same estimated values of \(\lambda\) may be
obtained from datasets containing vastly different input levels of
phylogenetic signal. Thus, \(\lambda\) is not a reliable estimate of the
strength of phylogenetic signal in phenotypic data. We subsequently
derive a standardized effect size for measuring the strength of
phylogenetic signal in phenotypic datasets, and apply the concept to two
common measures of phylogenetic signal: \(\lambda\) and \(Kappa\).
Through simulations across a wide range of conditions, we find that the
precision of effect sizes based on \(\lambda\) (\(Z_{\lambda}\)) are
less reliable than that those based on \(Kappa\) (\(Z_K\)), implying
that \(Z_K\) is a more robust effect size measure. Additionally, we
propose a two-sample test statistic that may be used to compare the
strength of phylogenetic signal among datasets, and provide an empirical
example to demonstrate its use. We conclude that estimates of
phylogenetic signal using Pagel's \(\lambda\) are often inaccurate, and
thus interpreting strength of phylogenetic signal in phenotypic datasets
based on this measure is compromised. By contrast, effect sizes obtained
from \(Kappa\) hold promise for characterizing phylogenetic signal, and
for comparing the strength of phylogenetic signal across datasets.

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal, with the worst
precision at intermediate levels of signal (\(\lambda_{in}\approx0.5\)),
and improved precision as input levels approached the extremes of its
range (i.e., \(\lambda_{in}\rightarrow0\) \&
\(\lambda_{in}\rightarrow1\)). Thus, estimates of \(\lambda\) were least
reflective of the true input signal at intermediate values.
Additionally, even at large levels of species richness, we found that
the range of \(\lambda_{est}\) still encompassed a substantial portion
of possible values (e.g., \(n=512\); \(\lambda_{in}=0.5\):
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\);
\(\lambda_{in} = 0.25 \to 0.65\)). Note that these findings were
particularly unsettling when considered in light of our literature
survey. Over one quarter of the \(\lambda\) estimates obtained in
empirical studies (421 of 1,572) were between \(\lambda=0.25\) and
\(\lambda=0.75\) (Figure 1). Of those, 30\% were interpreted directly in
terms of the strength of phylogenetic signal that they represented
(e.g., small, intermediate, strong signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression, the results of our simulations were quite similar.
Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). Additionally, estimates of phylogenetic signal were less
precise, and the spread of \(\lambda_{est}\) was similar to that
observed when \(\lambda\) was estimated for only the dependent variable
(Figure 3). Taken together, these findings reveal that \(\lambda_{est}\)
does not precisely characterize observed levels of phylogenetic signal
in phenotypic datasets, and that biological interpretations of the
strength of phylogenetic signal based on \(\lambda\) may be highly
inaccurate. \hfill\break

@@

Adams and Collyer (2019b) recently suggested that the strength of
phylogenetic signal could be represented as an effect size, based on the
\(Kappa\) statistic and its empirical sampling distribution from
permutation. Here we formalize that suggestion, resulting in an effect
size of:

@@

In this case, \(\lambda_{obs}\) and \(\hat\sigma_\lambda\) are
empirically derived using maximum likelihood. Note that under the null
hypothesis no phylogenetic signal is expected (Freckleton et al. 2002),
and thus \(E(\lambda)=0\) under this condition. \hfill\break

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4. Here two
things are evident. First, estimates of \(Z_K\) track the input
phylogenetic signal in a more linear fashion than do estimates of
\(Z_\lambda\) (Figure 4A,B). Thus, actual changes in the strength of
phylogenetic signal are reflected more evenly in the corresponding
values of the effect size \(Z_K\). Second, the precision of \(Z_K\) is
considerably more stable as compared with \(Z_\lambda\). This may be
seen by calculating the coefficients of variation for the set of
precision estimates (i.e., \(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)) across input levels of phylogenetic signal.
Here coefficients of variation in the precision of \(Z_K\) were an order
of magnitude smaller for than for \(Z_\lambda\) (Figure 4C,D). This
implied that estimates of the strength of phylogenetic signal were more
reliable and robust when using \(Z_K\) as compared with \(Z_\lambda\).
\hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), one may wish to compare such measures across datasets, to
determine whether the strength of phylogenetic signal is significantly
greater in one phenotypic trait as compared with another. As with other
effect sizes derived from permutation distributions (e.g., Adams and
Collyer 2016, 2019a), a two-sample test statistic may be found as:

@@

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above for
equation 2. Estimates of significance of \(\hat{Z}_{12}\) may be
obtained from a standard normal distribution. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (see Adams and Collyer 2016, 2019a). \hfill\break

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_K\)) are dimensionless, as the units of
measurement cancel out during their calculation {[}\textbf{S\&R??}{]}.
As a consequence, the effect sizes represent the strength of
phylogenetic signal on a common and comparable scale, measured in
standard deviation units, regardless of the initial units and scale of
the original phenotypic variables under investigation.

\hypertarget{section}{%
\subsubsection{}\label{section}}

Conceptually, this is akin to the conversion of data to their standard
normal deviates prior to their use in multivariate analysis, when the
original variables are expressed in different units or scales {[}see
{]}. \#\#\#

\hypertarget{section-1}{%
\paragraph{}\label{section-1}}

mathematical transformation or standardization of the trait dimensions,
so the variables are expressed in similar units n the variables in a
multivariate data set are on different scales,

Everitt, Brian; Hothorn, Torsten J (2011), An Introduction to Applied
Multivariate Analysis with R, Springer,

Also, Legendre\ldots{} Adams and Collyer. \#\#\#\#

Because of this standardization, the strength of phylogenetic signal
across datasets has meaning

something about comparisons having meaning\ldots{}

As such, characterizing the strength of phylogenetic signal with \(Z_K\)
facilitates a formal statistical comparison of values across continuous
traits of different units, scale, or even dimensionality (e.g.,
univariate traits using \(Kappa\), and multivariate traits using
\(K_{mult}\)).
